---
title: "CHOIR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CHOIR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Installation

To install the package, please use the following:
```{r, eval = FALSE}
remotes::install_github("corceslab/CHOIR", ref="main", repos = BiocManager::repositories(), upgrade = "never")
```

------

# Introduction

This vignette provides a basic example of how to run CHOIR, a clustering algorithm for single-cell sequencing data. CHOIR is applicable to single-cell sequencing data of any modality, including RNA, ATAC, and proteomics. It is also applicable to multi-modal data (see [Advanced Options](https://www.choirclustering.com/articles/CHOIR.html#advanced-options)). Detailed parameter definitions are available under the [Functions](https://www.choirclustering.com/reference/index.html) tab.

CHOIR is based on the premise that if clusters contain biologically different cell types or states, a machine learning classifier that considers features present in cells from each cluster should be able to distinguish the clusters with a higher level of accuracy than machine learning classifiers trained on randomly permuted cluster labels. The use of permutation testing approaches allows CHOIR to introduce statistical significance thresholds into the clustering process.

CHOIR proceeds in two main steps. First, CHOIR generates a hierarchical clustering tree spanning the network structure of the data from an initial “partition” in which all cells are in the same cluster to a partition in which all cells are demonstrably overclustered. Then, CHOIR applies a permutation testing approach using random forest classifiers across the nodes of the hierarchical clustering tree to determine the appropriate extent of each branch of the clustering tree, such that the final clusters represent statistically distinct populations.

You’ll need the following packages installed to run this tutorial:
```{r, message = FALSE, warning = FALSE}
library(CHOIR)
library(Seurat)
library(scRNAseq)
```

### Example data

To demonstrate how to run CHOIR on a single-cell RNA-seq dataset, we’ll use a small dataset consisting of mouse dopaminergic neurons, originating from an experiment by La Manno et al. (2016), and available through the Bioconductor package `scRNAseq`.

```{r, message = FALSE, warning = FALSE}
data_matrix <- LaMannoBrainData('mouse-adult')@assays@data$counts
colnames(data_matrix) <- seq(1:ncol(data_matrix))
```

### Pre-processing

CHOIR takes as input a Seurat, SingleCellExperiment, or ArchR object containing one or more feature matrices. Input data should already have undergone appropriate quality control and normalization steps (unless you are using SCTransform, see [Advanced Options](https://www.choirclustering.com/articles/CHOIR.html#advanced-options)). Note that the exact pre-processing steps used are up to the user.

This tutorial uses a Seurat object; for details related to other object types, see [Advanced Options](https://www.choirclustering.com/articles/CHOIR.html#advanced-options). First, we will create a Seurat object using the read count matrix. For simplicity, here we'll just exclude cells with fewer than 100 reads and genes present in less than 5 cells.

```{r, message = FALSE, warning = FALSE, results = FALSE}
seurat_object <- CreateSeuratObject(data_matrix, 
                                    min.features = 100,
                                    min.cells = 5)
```

We will now run log normalization.
```{r, message = FALSE, warning = FALSE, results = FALSE}
seurat_object <- NormalizeData(seurat_object)
```

------

# Running CHOIR

CHOIR proceeds in two main steps: 

1. CHOIR generates a hierarchical clustering tree.
2. CHOIR prunes this tree through the iterative application of a permutation testing approach.

The two steps can be run together using the function `CHOIR()` or separately using functions `buildTree()` and `pruneTree()`.

## Quick start:

The `CHOIR()` function will run all of the steps of the CHOIR algorithm in sequence. CHOIR is highly parallelized, so efficiency greatly improves as `n_cores` is increased.

The default significance level used by CHOIR is $\alpha = 0.05$ with Bonferroni multiple comparison correction. Other correction methods may be less conservative, as CHOIR applies filters that reduce the total number of tests performed (see [Advanced Options](https://www.choirclustering.com/articles/CHOIR.html#advanced-options)). 

We recommend using the default value of $\alpha = 0.05$ with Bonferroni multiple comparison correction. For a more conservative approach, the `alpha` value could be decreased to 0.01 or 0.001.

```{r, eval = FALSE}
seurat_object <- CHOIR(seurat_object, 
                       n_cores = 2)
```

## Step-by-step:

Alternately, we can run each of the two main steps of CHOIR individually. 

For users who already have a set of clusters generated by a different tool, and wish to assess whether these clusters are under- or over-clustered, start at Step 2 (see [Advanced Options](https://www.choirclustering.com/articles/CHOIR.html#advanced-options)).

### Step 1: Generate hierarchical clustering tree
CHOIR generates a hierarchical clustering tree by computing dimensionality reduction and nearest neighbor adjacency matrices for the data, computing clusters across a series of sampled resolutions, then reconciling those clustering results into a consensus clustering tree using package `MRtree`. When using the default parameter `max_clusters = 'auto'`, CHOIR applies a series of subclustering steps to ensure that the tree is not underclustered. Specifically, the silhouette score is assessed at each new level of the emerging tree. When the silhouette score is maximized, each cluster at this level is subset, a new dimensionality reduction is generated, and the cluster is then subclustered. Each of these subtrees continues to be subdivided until the farthest pair of nearest neighboring clusters is found to be overclustered using the permutation test approach.

The user can determine which method is used for dimensionality reduction and batch correction, as well as supply specific parameters for dimensionality reduction, batch correction, detecting nearest neighbors, and modularity-based clustering, as needed. 

Here, we will build the hierarchical clustering tree using default parameter settings.

```{r, message = TRUE, warning = FALSE, results = "hide"}
seurat_object <- buildTree(seurat_object,
                           n_cores = 2)
```

### Step 2: Prune hierarchical clustering tree
After constructing the hierarchical clustering tree, CHOIR iterates through each node of the clustering tree using a bottom-up approach. At each node, CHOIR computes pairwise comparisons of all child clusters by splitting cells into training and test sets, training a balanced random forest classifier on the gene expression data of the training set, and then predicting the cluster assignments of the cells in the test set. This yields a prediction accuracy score which represents the degree to which the two clusters are distinguishable.

In parallel, CHOIR shuffles the cluster labels and repeats the same process. Both comparisons are repeated using bootstrapped samples (default = 100 iterations), resulting in a permutation test that compares the true prediction accuracy for the clusters to the prediction accuracy for a chance division of the cells into two random groups.

This permutation test yields a p-value that determines whether these clusters are slated to merge or remain separate. The significance threshold used can be adjusted using the `alpha` parameter. We recommend using the default value of $\alpha = 0.05$ with Bonferroni multiple comparison correction. For a more conservative approach, the `alpha` value could be decreased to 0.01 or 0.001.

```{r, message = TRUE, warning = FALSE, results = "hide"}
seurat_object <- pruneTree(seurat_object, 
                           n_cores = 2)
```

## Plot

Labels for the final clusters identified by CHOIR can be found in the `meta.data` slot of the Seurat object. Other CHOIR outputs are stored under the `misc` slot of the Seurat object.

```{r}
head(seurat_object@meta.data)
```

After clustering has finished, we can generate a UMAP dimensionality reduction and use function `plotCHOIR()` to plot a visualization of the clusters.

```{r}
# Run UMAP
seurat_object <- runCHOIRumap(seurat_object,
                              reduction = "P0_reduction")
# Plot
plotCHOIR(seurat_object)
```

This function will also overlay the prediction accuracy scores among neighboring pairs of clusters when `accuracy_scores` is set to `TRUE`.

```{r}
plotCHOIR(seurat_object,
          accuracy_scores = TRUE,
          plot_nearest = FALSE)
```

------

# Advanced Options

## Input object types

### Seurat

For Seurat objects, the default assay is used if no input is provided for parameter `use_assay`. If no input is provided for parameter `use_slot`, the following default slot (Seurat v4) or layer (Seurat v5) is used:

*  If `use_assay` is either "RNA" or "sketch", CHOIR will look for the `data` slot
*  If `use_assay` is either "SCT" or "integrated", CHOIR will look for the `scale.data` slot

The default dimensionality reduction method for Seurat objects is 'PCA', except in the case of ATAC-seq data, where it is 'LSI'.

If you would like to use SCTransform normalization rather than log normalization, please provide raw counts and set the parameter `normalization_method` to 'SCTransform'. Note that SCTransform has not been thoroughly tested with CHOIR.

Labels for the final clusters identified by CHOIR can be found in the `meta.data` slot of the Seurat object. Other CHOIR outputs are stored under the `misc` slot of the Seurat object.

### SingleCellExperiment

For SingleCellExperiment objects, only the `use_assay` parameter is needed. If not provided, it is set to 'logcounts'.

The default dimensionality reduction method for SingleCellExperiment objects is 'PCA', except in the case of ATAC-seq data, where it is 'LSI'.

Labels for the final clusters identified by CHOIR can be found in the `colData` slot of the SingleCellExperiment object. Other CHOIR outputs are stored under `metadata`.

### ArchR

For ArchR objects, if no input is provided for parameter `ArchR_matrix`, the "TileMatrix" is used. If no input for parameter `ArchR_depthcol` is provided, "nFrags" is used.

The default dimensionality reduction method for ArchR objects is 'IterativeLSI'.

Labels for the final clusters identified by CHOIR can be found in the `cellColData` slot of the ArchR object. Other CHOIR outputs are stored under `projectMetadata`.

## CHOIR parameters

### Batch correction

For datasets with multiple batches, it is recommended to apply Harmony batch correction through CHOIR by setting the parameter `batch_correction_method` to 'Harmony'. This not only generates Harmony-corrected dimensionality reductions, but ensures that random forest classifer comparisons are batch-aware. 

Use caution in applying this method if your groups of interest (e.g., disease vs. control) are batch-confounded AND you expect cell types unique to each of these groups.

### Significance level & multiple comparison correction

The default significance level used by CHOIR is $\alpha = 0.05$ with Bonferroni multiple comparison correction. Other correction methods may be less conservative, as `CHOIR` applies filters that reduce the total number of tests performed (see below). 

We recommend using the default value of $\alpha = 0.05$ with Bonferroni multiple comparison correction. For a more conservative approach, the `alpha` value could be decreased to 0.01 or 0.001.

### Filters

CHOIR uses various filters to reduce the number of necessary permutation test comparisons:

*  Parameter `min_accuracy` indicates the minimum accuracy for the random forest classifier predictions, below which clusters will be automatically merged. It defaults to 0.5, higher values will identify a more conservative set of clusters.
*  Parameter `min_connections` indicates the minimum number of nearest neighbors between two clusters for them to be considered 'adjacent'. If non-zero, non-adjacent clusters will never be merged. Defaults to 1.
*  Parameter `max_repeat_errors` is used to account for situations in which random forest classifier prediction errors are concentrated among a few cells that are repeatedly misassigned (often doublets or other outliers). If non-zero, repeated errors will be evaluated and accuracy scores will be modified if the number of repeated errors is below the value of `max_repeat_errors`. Defaults to 20.
*  Parameter `distance_awareness` represents the distance threshold above which a cluster will not merge with another cluster. Specifically, this value is a multiplier applied to the distance between a cluster and its closest distinguishable neighbor; the product sets the threshold for subsequent comparisons. The default value sets this threshold at a 2-fold increase in distance.

### Downsampling

CHOIR uses downsampling to increase efficiency for larger datasets. Using the default parameter setting of `downsampling_rate = "auto"`, downsampling occurs at each random forest classifer comparison. The downsampling rate is determined based on the overall dataset size. To disable downsampling, set `downsampling_rate = 1`.

Additional downsampling can be imposed using parameter `sample_max`, indicating the maximum number of cells used per cluster to train/test each random forest classifier. By default, this is not used.

## Providing pre-generated clusters

For users who already have a set of clusters generated by a different tool, and wish to assess whether these clusters are under- or over-clustered, only need the `pruneTree()` function.

To `pruneTree()`, provide:

* `object` The input object under which the results will be stored.
* `cluster_tree` A dataframe containing the cluster IDs of each cell across the levels of a hierarchical clustering tree.
* `input_matrix` A matrix containing the feature x cell data on which to train the random forest classifiers. 
* `nn_matrix` A matrix containing the nearest neighbor adjacency of the cells. 
* Either reduction (a matrix of dimensionality reduction cell embeddings) if using approximate distances OR `dist_matrix` (a distance matrix of cell to cell distances)

## Using CHOIR with ATAC-seq data

If your input data is ATAC-seq data, make sure to set input parameter `atac` to `TRUE`, in order to use the correct defaults. For example, CHOIR uses 25000 variable features by default for ATAC-seq data, in comparison to 2000 by default for other data. If your data exclusively contains ATAC-seq data (no other modalities), it is recommended to set parameter `use_variance` to `FALSE`.

## Using CHOIR with multi-modal data

If you are providing input data which contains multiple modalities from the same cell, such as RNA + ATAC multiome data, you must provide multiple values (as a vector) to parameters `use_assay` and `use_slot` (for Seurat objects), `use_assay` for (SingleCellExperiment objects), or `ArchR_matrix`and `ArchR_depthcol` for ArchR objects.

If you want to use different methods of normalization, dimensionality reduction, number of variable features, or batch correction for each modality, input to related parameters must also be provided as vectors or lists with one value per modality.

CHOIR creates a joint dimensionality reduction for the modalities and uses all provided feature matrices as joint input to train/test random forest classifiers.
